Policing for Profit and the Undermining of Democracy\thanks{The authors thank XX for their feedback and support. All errors are our responsibility.}

### 2021-01-27 markdown/fees_fines_to.Rmd
add 2way fes

---
title: "Policing for Profit and the Undermining of Democracy"
subtitle: "Fines, Fees, and Turnout"
thanks: The authors thank XX for their feedback and support. All errors are our responsibility. 
author:
- Jonathan Ben-Menachem^[PhD Student, Columbia University, Department of Sociology (jbenmenachem@gmail.com)]
- Kevin Morris^[Researcher, Brennan Center for Justice at NYU School of Law (kevin.morris@nyu.edu)]
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  bookdown::pdf_document2:
    toc: no
    fig_caption: yes
    latex_engine: xelatex
    keep_tex: true
bibliography: "policing_turnout.bib"
link-citations: yes
fontsize: 12pt
header-includes:
    - \usepackage{rotating}
    - \usepackage{setspace}
abstract: "This is an abstract."
---
\pagenumbering{gobble}
\pagebreak
```{r setup, echo = F, message = F, warning = F}
knitr::opts_knit$set(root.dir = "..", dev = "cairo_pdf")
library(systemfit)
library(kevostools)
library(AER)
library(sandwich)
library(lmtest)
library(scales)
library(ggpubr)
library(tidyverse)
library(extrafont)
library(kableExtra)
library(data.table)
library(stargazer)
loadfonts()
```
\pagenumbering{arabic}
\begin{center}
Policing for Profit and the Undermining of Democracy: Fines, Fees, and Turnout \\
Jonathan Ben-Menachem and Kevin Morris
\end{center}

Fines and fees are increasingly recognized as a form of racist revenue extraction connected to marginalized communities’ alienation from government. After Michael Brown was killed by the Ferguson Police Department in 2014, a US Department of Justice investigation into the city’s police and courts demonstrated that the municipality was engaged in a practice that advocates now refer to as “policing for profit.” The city’s reliance on fines and fees to fund government functions grew from 13 to 23 percent of the total budget between fiscal years 2012 and 2015 [@UnitedStatesDepartmentofJusticeCivilRightsDivision2015].

It wasn’t just a Ferguson problem, or even a Missouri problem. American cities’ reliance on fines and fees revenue increased significantly following the 2008 recession—as local tax revenues dropped and tax increases became less politically viable, jurisdictions increased the amounts of fines and fees and imposed them more frequently in order to fund government services [@Singla2020; @Mughan2020; @Kirk2020; @Martin2018]. Although a body of literature developed over the past decade indicates that criminalization chills political participation, no work has directly investigated the relationship between fines and fees revenue collection and voter turnout. This project intends to fill that gap.

@@

@@

Ticketing practices may be related to distrust of government and reduced political participation. In a survey of residents in three Georgia cities that rely heavily on fines and fees revenue, residents who were ticketed reported lower levels of trust in police and government more broadly [@CarpenterII2019]. Lerman and Weaver have argued that residents’ perceptions of law enforcement may affect their perceptions of government as a whole. As one man told them, “I feel like if I contact a senator or governor, they’ll probably want to put me in jail and leave me as a troublemaker... I try to basically stay away from government” [-@Lerman2014, p. 210]. In their study of descriptive representation, Sances and You [-@Sances2017] explicitly suggested that “fines may make descriptive representation less likely by depressing minority turnout,” and called for further research regarding voter turnout and the imposition of fines. In this paper, we clarify that relationship.

We begin by picking up where @Sances2017 left off. By tying data from the Census of Governments to a national geocoded voter file we test whether municipalities with more fees and fines collected per resident saw lower turnout in the 2014, 2016, and 2018 general elections. We expect to find that municipalities where more revenue is raised per resident saw lower turnout, other things being equal. Because the American criminal legal system falls disproportionately on minorities, we also expect that the racial turnout gap [@Fraga2018] will be larger in municipalities that rely more heavily on fines and fees.

Of course, fines and fees imposition may be related to turnout without necessarily indicating a causal relationship. To estimate the causal effect, we will explore whether turnout increases or decreases in the wake of changes in city revenue collection policy.

@@

In order to ascertain whether fines and fees practices affect voter turnout, we exploit data from the 2013 and 2017 Census of Governments (COG) and national voter file data. The COG, a project of the US Census Bureau, collects budget data for all local governments every five years, and the most recent survey was in 2017. The COG asks cities how much revenue they collect from “penalties imposed for violation of law; civil penalties (e.g., for violating court orders); court fees if levied upon conviction of a crime or violation... and forfeits of deposits held for performance guarantees or against loss or damage (such as forfeited bail and collateral).” As indicated by previous researchers, these data serve as a useful proxy for the prevalence of police ticketing [@Goldstein2020; @Sances2017].

Unfortunately, the COG data does not include voter turnout, and turnout data is not generally available at this level. Therefore, in order to determine municipality-level turnout, we employ a national, geocoded voter file. This voter file is provided by L2 Political, and includes a host of information about each voter, including age, party affiliation, racial estimates, and historical turnout. By using the latitude and longitude of every registered voter in the country and shapefiles provided by the Census Bureau, we map each voter to their home municipality using the same definitions as the COG.^[Although the registered voter file includes the name of the city in which the voter is registered, these do not always map neatly to the municipalities defined in the COG data. The use of these shapefiles ensures an apples-to-apples comparison.] We then aggregate these individual-level records up to the municipality level to determine the number of ballots cast in each municipality. TUrnout is calculated by dividing this number by the city’s citizen voting age population (CVAP) from the Census Bureau.

Because the L2 national voter file includes estimates of each voter’s racioethnic identity, we can also use these data to determine the *racial turnout gap* in each municipality. These records are just estimates, and may be imprecise at the individual level. However, we assume that this presents less of a problem when aggregating to the municipality-level. In order to control for potential confounders, we also control for a host of other information from the 5-year American Community Survey.

Although we do not present these results in this abstract, we also intend to incorporate causal estimates in the full version of this study by using changes in fees and fines reliance between the 2013 and 2017 COGs, and turnout in the 2014 and 2018 elections.

@@

Figure \ref{fig:marg1} makes clear that there is a relationship between the fees and fines collected by a municipality and its turnout. The left-hand panel shows the simple bivariate relationship between the two quantities. In the right-hand panel, we present the marginal effects plot in lieu of a regression table due to space constraints. The second panel shows that, even after controlling for a municipality’s racial composition, median income, collegiate education, age, and state fixed effects, there is a statistically significant, negative relationship between fees and fines and turnout in 2018. In these models, the dependent variable is municipal turnout in 2018 and the independent variable of interest is the natural log of the dollars in fees and fines per resident plus 1.

```{r, figures-side, fig.show="hold", out.width="50%", echo = FALSE, message = F, warning = F, fig.cap = "\\label{fig:marg1}Dollars per Resident and 2018 Turnout", fig.pos = "H"}
p1 <- readRDS("./temp/simple_scat.rds")
load("./temp/mef_to.rdata")
print(p1)
print(p2)
```

All else equal, a 1 percent increase in the dollars of fees and fines raised per resident is associated with a reduction in turnout of 0.003 percentage points, and the relationship is significant at the 99 percent confidence level. Although this may seem small, Figure \ref{fig:marg1} shows that the range of fees and fines raised is large. The figure marks the 10th and 90th percentiles for revenue collected in fees and fines, and predicted turnout drops by 0.8 percentage points over this range.

Figure \ref{fig:marg2} presents the results of preliminary analyses investigating the relationship between fees and fines collection and the white-Black turnout gap. Racial turnout figures are calculated by using the number of ballots cast by members of each racial group according to L2 divided by the CVAP estimates for that group in each municipality. Our initial analysis indicates that the white-Black turnout gap in a municipality is strongly associated with the level of fees and fines it imposes. Figure \ref{fig:marg2} shows that the racial turnout gap is much larger in cities where there are fewer fees and fines imposed. Moving from the 10th to 90th percentile of fees and fines per capita is associated with a 9 percentage point drop in the white-Black turnout gap.

```{r, fig.show="hold", out.width="50%", echo = FALSE, message = F, warning = F, fig.cap = "\\label{fig:marg2}Dollars per Resident and 2018 Turnout Gap", fig.pos = "H"}
p5 <- readRDS("./temp/simple_scat_gap_bw.rds")
load("./temp/mef_gap_bw.rdata")
print(p5)
print(p6)
```

>>>>

---
title: "Policing for Profit and the Undermining of Democracy"
subtitle: "Fines, Fees, and Turnout"
thanks: The authors thank XX for their feedback and support. All errors are our responsibility. 
author:
- Jonathan Ben-Menachem^[PhD Student, Columbia University, Department of Sociology (jb4487@columbia.edu)]
- Kevin Morris^[Researcher, Brennan Center for Justice at NYU School of Law (kevin.morris@nyu.edu)]
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  bookdown::pdf_document2:
    toc: no
    fig_caption: yes
    latex_engine: xelatex
    keep_tex: true
bibliography: "policing_turnout.bib"
link-citations: yes
fontsize: 12pt
header-includes:
    - \usepackage{rotating}
    - \usepackage{setspace}
abstract: "This is an abstract."
---
\pagenumbering{gobble}
\pagebreak
```{r setup, echo = F, message = F, warning = F}
knitr::opts_knit$set(root.dir = "..", dev = "cairo_pdf")
library(systemfit)
library(kevostools)
library(AER)
library(sandwich)
library(lmtest)
library(scales)
library(ggpubr)
library(tidyverse)
library(extrafont)
library(kableExtra)
library(data.table)
library(stargazer)
loadfonts()
```
\pagenumbering{arabic}
\begin{center}
Policing for Profit and the Undermining of Democracy: Fines, Fees, and Turnout \\
Jonathan Ben-Menachem and Kevin Morris
\end{center}

Fines and fees are increasingly recognized as a form of racist revenue extraction connected to marginalized communities’ alienation from government. American cities’ reliance on fines and fees revenue increased significantly following the 2008 recession—as local tax revenues dropped and tax increases became less politically viable, jurisdictions increased the amounts of fines and fees and imposed them more frequently in order to fund government services [@Singla2020; @Mughan2020; @Kirk2020; @Martin2018]. Although a body of literature developed over the past decade indicates that criminalization chills political participation, no work has directly investigated the relationship between fines and fees revenue collection and voter turnout. This project intends to fill that gap by testing whether fines and fees at the municipal level influence voter turnout.

@@

Additionally, Sances and You [-@Sances2017] explicitly suggested that “fines may make descriptive representation less likely by depressing minority turnout,” and called for further research regarding voter turnout and the imposition of fines. In this paper, we clarify that relationship by picking up where @Sances2017 left off. By tying data from the 2012 and 2017 Census of Governments to a national geocoded voter file we test whether municipalities with more fees and fines collected per resident saw lower turnout in the 2014 and 2018 general elections. We expect to find that municipalities where more revenue is raised per resident saw lower turnout, other things being equal. Because the American criminal legal system falls disproportionately on minorities, we also expect that the racial turnout gap [@Fraga2018] will be larger in municipalities that rely more heavily on fines and fees. Of course, fines and fees imposition may be related to turnout without necessarily indicating a causal relationship. For this reason we leverage data on fines collections and turnout from multiple time periods to investigate the causal relationship.

@@

Public perceptions of abusive police practices, such as high-profile incidents of police violence or racially discriminatory street stops, can reduce willingness to report crimes or cooperate with law enforcement [@Desmond2016; @Tyler2014). Legitimacy studies have largely focused on police violence or police street stops, but new evidence related to fines and fees has recently emerged. 
In a survey of residents in three Georgia cities that rely heavily on fines and fees revenue, residents who were ticketed reported lower levels of trust in police and government more broadly [@CarpenterII2019].

Although the conceptual framework for police legitimacy has been fruitfully critiqued—Monica @Bell2017 points to legal estrangement and structural exclusion in order to shift the theoretical focus from perceptions of unfairness to substantively unjust practices—the empirical literature on legitimacy is useful to demonstrate how police practices can alienate those subjected to them.  The alienating effect of criminal legal exposure has also been tied to institutional avoidance [see, for instance, @Brayne2014]. .

@@

In order to ascertain whether fines and fees practices affect voter turnout, we exploit data from the 2012 and 2017 Census of Governments (COG) and national voter file data. The COG, a project of the US Census Bureau, collects budget data for all local governments every five years, and the most recent survey was in 2017. The COG asks cities how much revenue they collect from “penalties imposed for violation of law; civil penalties (e.g., for violating court orders); court fees if levied upon conviction of a crime or violation... and forfeits of deposits held for performance guarantees or against loss or damage (such as forfeited bail and collateral).” As indicated by previous researchers, these data serve as a useful proxy for the prevalence of police ticketing [@Goldstein2020; @Sances2017].

Unfortunately, the COG data does not include voter turnout, and turnout data is not generally available at this level. Therefore, in order to determine municipality-level turnout, we employ a national, geocoded voter file. This voter file is provided by L2 Political, and includes a host of information about each voter, including age, party affiliation, racial estimates, and historical turnout. By using the latitude and longitude of every registered voter in the country and shapefiles provided by the Census Bureau, we map each voter to their home municipality using the same definitions as the COG.^[Although the registered voter file includes the name of the city in which the voter is registered, these do not always map neatly to the municipalities defined in the COG data. The use of these shapefiles ensures an apples-to-apples comparison.] We then aggregate these individual-level records up to the municipality level to determine each municipality’s turnout.

Because the L2 national voter file includes estimates of each voter’s racioethnic identity, we can also use these data to determine the *racial turnout gap* in each municipality. These records are just estimates, and may be imprecise at the individual level. However, we assume that this presents less of a problem when aggregating to the municipality-level. In order to control for potential confounders, we also control for a host of other information from the 5-year American Community Survey and the COG.

@@

Figure \ref{fig:marg1} makes clear that higher fees and fines collections lead to lower turnout. The left-hand panel shows the simple bivariate relationship between turnout among registered voters in 2018 and fees and fines collection. The right-hand panel shows the bivariate relationship between the 2018 white-Black turnout gap and fines collections. The turnout gap is calculated by dividing the number of ballots cast by each racial group according to L2 by the estimated 2018 citizen voting age population (CVAP) from the Census Bureau to account for possible registration differences between groups.

```{r, figures-side, fig.show="hold", out.width="50%", echo = FALSE, message = F, warning = F, fig.cap = "\\label{fig:marg1}Dollars per Resident and 2018 Turnout", fig.pos = "H"}
p1 <- readRDS("./temp/simple_scat.rds")
p2 <- readRDS("./temp/simple_scat_gap_bw.rds")
print(p1)
print(p2)
```

Although Figure \ref{fig:marg1} shows that there is a relationship between collections and turnout and the turnout gap, the simple scatter plots cannot demonstrate a causal relationship. To investigate whether there is such a relationship, we employ two-way fixed effect models. In these models, each municipality is observed twice: their first observation includes their (logged + 1) fees and fines per capita from the 2012 COG, their turnout and turnout gap in 2014, and their sociodemographic characteristics from the ACS 5-year estimates from 2012. The second observation includes their (logged + 1) fees and fines from the 2017 COG, their 2018 turnout (gap), and their 2018 5-year ACS demographics. In each case, turnout is measured only among the individuals who were registered in a given municipality in both 2014 and 2018 to control for changes in registration rates, and in Model 2 the turnout gap is calculated using CVAP. Table \ref{tab:tab1} presents the results of these models.

\begin{singlespace}
\input{"../temp/2wfe_reg_clean.tex"}
\end{singlespace}

### 2021-01-27 markdown/fees_fines_to.tex
add 2way fes

\title{Policing for Profit and the Undermining of Democracy\thanks{The authors thank XX for their feedback and support. All errors are our responsibility.}}
\usepackage{etoolbox}
\makeatletter
\providecommand{\subtitle}[1]{% add subtitle to \maketitle
  \apptocmd{\@title}{\par {\large #1 \par}}{}{}
}
\makeatother
\subtitle{Fines, Fees, and Turnout}
\author{Jonathan Ben-Menachem\footnote{PhD Student, Columbia University, Department of Sociology (\href{mailto:jbenmenachem@gmail.com}{\nolinkurl{jbenmenachem@gmail.com}})} \and Kevin Morris\footnote{Researcher, Brennan Center for Justice at NYU School of Law (\href{mailto:kevin.morris@nyu.edu}{\nolinkurl{kevin.morris@nyu.edu}})}}
\date{January 14, 2021}

@@

Fines and fees are increasingly recognized as a form of racist revenue extraction connected to marginalized communities' alienation from government. After Michael Brown was killed by the Ferguson Police Department in 2014, a US Department of Justice investigation into the city's police and courts demonstrated that the municipality was engaged in a practice that advocates now refer to as ``policing for profit.'' The city's reliance on fines and fees to fund government functions grew from 13 to 23 percent of the total budget between fiscal years 2012 and 2015 (United States Department of Justice Civil Rights Division \protect\hyperlink{ref-UnitedStatesDepartmentofJusticeCivilRightsDivision2015}{2015}).

It wasn't just a Ferguson problem, or even a Missouri problem. American cities' reliance on fines and fees revenue increased significantly following the 2008 recession---as local tax revenues dropped and tax increases became less politically viable, jurisdictions increased the amounts of fines and fees and imposed them more frequently in order to fund government services (Singla, Kirschner, and Stone \protect\hyperlink{ref-Singla2020}{2020}; Mughan \protect\hyperlink{ref-Mughan2020}{2020}; Kirk, Fernandes, and Friedman \protect\hyperlink{ref-Kirk2020}{2020}; Martin \protect\hyperlink{ref-Martin2018}{2018}). Although a body of literature developed over the past decade indicates that criminalization chills political participation, no work has directly investigated the relationship between fines and fees revenue collection and voter turnout. This project intends to fill that gap.

@@

@@

Ticketing practices may be related to distrust of government and reduced political participation. In a survey of residents in three Georgia cities that rely heavily on fines and fees revenue, residents who were ticketed reported lower levels of trust in police and government more broadly (Carpenter II, Sweetland, and McDonald \protect\hyperlink{ref-CarpenterII2019}{2019}). Lerman and Weaver have argued that residents' perceptions of law enforcement may affect their perceptions of government as a whole. As one man told them, ``I feel like if I contact a senator or governor, they'll probably want to put me in jail and leave me as a troublemaker\ldots{} I try to basically stay away from government'' (\protect\hyperlink{ref-Lerman2014}{2014}, 210). In their study of descriptive representation, Sances and You (\protect\hyperlink{ref-Sances2017}{2017}) explicitly suggested that ``fines may make descriptive representation less likely by depressing minority turnout,'' and called for further research regarding voter turnout and the imposition of fines. In this paper, we clarify that relationship.

We begin by picking up where Sances and You (\protect\hyperlink{ref-Sances2017}{2017}) left off. By tying data from the Census of Governments to a national geocoded voter file we test whether municipalities with more fees and fines collected per resident saw lower turnout in the 2014, 2016, and 2018 general elections. We expect to find that municipalities where more revenue is raised per resident saw lower turnout, other things being equal. Because the American criminal legal system falls disproportionately on minorities, we also expect that the racial turnout gap (Fraga \protect\hyperlink{ref-Fraga2018}{2018}) will be larger in municipalities that rely more heavily on fines and fees.

Of course, fines and fees imposition may be related to turnout without necessarily indicating a causal relationship. To estimate the causal effect, we will explore whether turnout increases or decreases in the wake of changes in city revenue collection policy.

@@

In order to ascertain whether fines and fees practices affect voter turnout, we exploit data from the 2013 and 2017 Census of Governments (COG) and national voter file data. The COG, a project of the US Census Bureau, collects budget data for all local governments every five years, and the most recent survey was in 2017. The COG asks cities how much revenue they collect from ``penalties imposed for violation of law; civil penalties (e.g., for violating court orders); court fees if levied upon conviction of a crime or violation\ldots{} and forfeits of deposits held for performance guarantees or against loss or damage (such as forfeited bail and collateral).'' As indicated by previous researchers, these data serve as a useful proxy for the prevalence of police ticketing (Goldstein, Sances, and You \protect\hyperlink{ref-Goldstein2020}{2020}; Sances and You \protect\hyperlink{ref-Sances2017}{2017}).

Unfortunately, the COG data does not include voter turnout, and turnout data is not generally available at this level. Therefore, in order to determine municipality-level turnout, we employ a national, geocoded voter file. This voter file is provided by L2 Political, and includes a host of information about each voter, including age, party affiliation, racial estimates, and historical turnout. By using the latitude and longitude of every registered voter in the country and shapefiles provided by the Census Bureau, we map each voter to their home municipality using the same definitions as the COG.\footnote{Although the registered voter file includes the name of the city in which the voter is registered, these do not always map neatly to the municipalities defined in the COG data. The use of these shapefiles ensures an apples-to-apples comparison.} We then aggregate these individual-level records up to the municipality level to determine the number of ballots cast in each municipality. TUrnout is calculated by dividing this number by the city's citizen voting age population (CVAP) from the Census Bureau.

Because the L2 national voter file includes estimates of each voter's racioethnic identity, we can also use these data to determine the \emph{racial turnout gap} in each municipality. These records are just estimates, and may be imprecise at the individual level. However, we assume that this presents less of a problem when aggregating to the municipality-level. In order to control for potential confounders, we also control for a host of other information from the 5-year American Community Survey.

Although we do not present these results in this abstract, we also intend to incorporate causal estimates in the full version of this study by using changes in fees and fines reliance between the 2013 and 2017 COGs, and turnout in the 2014 and 2018 elections.

@@

Figure \ref{fig:marg1} makes clear that there is a relationship between the fees and fines collected by a municipality and its turnout. The left-hand panel shows the simple bivariate relationship between the two quantities. In the right-hand panel, we present the marginal effects plot in lieu of a regression table due to space constraints. The second panel shows that, even after controlling for a municipality's racial composition, median income, collegiate education, age, and state fixed effects, there is a statistically significant, negative relationship between fees and fines and turnout in 2018. In these models, the dependent variable is municipal turnout in 2018 and the independent variable of interest is the natural log of the dollars in fees and fines per resident plus 1.

@@

All else equal, a 1 percent increase in the dollars of fees and fines raised per resident is associated with a reduction in turnout of 0.003 percentage points, and the relationship is significant at the 99 percent confidence level. Although this may seem small, Figure \ref{fig:marg1} shows that the range of fees and fines raised is large. The figure marks the 10th and 90th percentiles for revenue collected in fees and fines, and predicted turnout drops by 0.8 percentage points over this range.

Figure \ref{fig:marg2} presents the results of preliminary analyses investigating the relationship between fees and fines collection and the white-Black turnout gap. Racial turnout figures are calculated by using the number of ballots cast by members of each racial group according to L2 divided by the CVAP estimates for that group in each municipality. Our initial analysis indicates that the white-Black turnout gap in a municipality is strongly associated with the level of fees and fines it imposes. Figure \ref{fig:marg2} shows that the racial turnout gap is much larger in cities where there are fewer fees and fines imposed. Moving from the 10th to 90th percentile of fees and fines per capita is associated with a 9 percentage point drop in the white-Black turnout gap.

\begin{figure}[H]
\includegraphics[width=0.5\linewidth]{fees_fines_to_files/figure-latex/unnamed-chunk-1-1} \includegraphics[width=0.5\linewidth]{fees_fines_to_files/figure-latex/unnamed-chunk-1-2} \caption{\label{fig:marg2}Dollars per Resident and 2018 Turnout Gap}\label{fig:unnamed-chunk-1}
\end{figure}

@@

\hypertarget{refs}{}
\begin{cslreferences}
\leavevmode\hypertarget{ref-CarpenterII2019}{}%
Carpenter II, Dick M, Kyle Sweetland, and Jennifer McDonald. 2019. ``The Price of Taxation by Citation: Case Studies of Three Georgia Cities That Rely Heavily on Fines and Fees.'' Institute for Justice.

@@

@@

\leavevmode\hypertarget{ref-Lerman2014}{}%
Lerman, Amy E., and Vesla M. Weaver. 2014. \emph{Arresting Citizenship: The Democratic Consequences of American Crime Control}. Chicago Studies in American Politics. Chicago ; London: The University of Chicago Press.

@@

@@

\leavevmode\hypertarget{ref-UnitedStatesDepartmentofJusticeCivilRightsDivision2015}{}%
United States Department of Justice Civil Rights Division. 2015. ``Investigation of the Ferguson Police Department.'' \url{https://www.justice.gov/sites/default/files/opa/press-releases/attachments/2015/03/04/ferguson_police_department_report.pdf}.

>>>>

\title{Policing for Profit and the Undermining of Democracy\thanks{The authors thank XX for their feedback and support. All errors are our responsibility.}}
\usepackage{etoolbox}
\makeatletter
\providecommand{\subtitle}[1]{% add subtitle to \maketitle
  \apptocmd{\@title}{\par {\large #1 \par}}{}{}
}
\makeatother
\subtitle{Fines, Fees, and Turnout}
\author{Jonathan Ben-Menachem\footnote{PhD Student, Columbia University, Department of Sociology (\href{mailto:jb4487@columbia.edu}{\nolinkurl{jb4487@columbia.edu}})} \and Kevin Morris\footnote{Researcher, Brennan Center for Justice at NYU School of Law (\href{mailto:kevin.morris@nyu.edu}{\nolinkurl{kevin.morris@nyu.edu}})}}
\date{January 27, 2021}

@@

Fines and fees are increasingly recognized as a form of racist revenue extraction connected to marginalized communities' alienation from government. American cities' reliance on fines and fees revenue increased significantly following the 2008 recession---as local tax revenues dropped and tax increases became less politically viable, jurisdictions increased the amounts of fines and fees and imposed them more frequently in order to fund government services (Singla, Kirschner, and Stone \protect\hyperlink{ref-Singla2020}{2020}; Mughan \protect\hyperlink{ref-Mughan2020}{2020}; Kirk, Fernandes, and Friedman \protect\hyperlink{ref-Kirk2020}{2020}; Martin \protect\hyperlink{ref-Martin2018}{2018}). Although a body of literature developed over the past decade indicates that criminalization chills political participation, no work has directly investigated the relationship between fines and fees revenue collection and voter turnout. This project intends to fill that gap by testing whether fines and fees at the municipal level influence voter turnout.

@@

Additionally, Sances and You (\protect\hyperlink{ref-Sances2017}{2017}) explicitly suggested that ``fines may make descriptive representation less likely by depressing minority turnout,'' and called for further research regarding voter turnout and the imposition of fines. In this paper, we clarify that relationship by picking up where Sances and You (\protect\hyperlink{ref-Sances2017}{2017}) left off. By tying data from the 2012 and 2017 Census of Governments to a national geocoded voter file we test whether municipalities with more fees and fines collected per resident saw lower turnout in the 2014 and 2018 general elections. We expect to find that municipalities where more revenue is raised per resident saw lower turnout, other things being equal. Because the American criminal legal system falls disproportionately on minorities, we also expect that the racial turnout gap (Fraga \protect\hyperlink{ref-Fraga2018}{2018}) will be larger in municipalities that rely more heavily on fines and fees. Of course, fines and fees imposition may be related to turnout without necessarily indicating a causal relationship. For this reason we leverage data on fines collections and turnout from multiple time periods to investigate the causal relationship.

@@

Public perceptions of abusive police practices, such as high-profile incidents of police violence or racially discriminatory street stops, can reduce willingness to report crimes or cooperate with law enforcement {[}Desmond, Papachristos, and Kirk (\protect\hyperlink{ref-Desmond2016}{2016}); Tyler, Fagan, and Geller (\protect\hyperlink{ref-Tyler2014}{2014})). Legitimacy studies have largely focused on police violence or police street stops, but new evidence related to fines and fees has recently emerged.
In a survey of residents in three Georgia cities that rely heavily on fines and fees revenue, residents who were ticketed reported lower levels of trust in police and government more broadly (Carpenter II, Sweetland, and McDonald \protect\hyperlink{ref-CarpenterII2019}{2019}).

Although the conceptual framework for police legitimacy has been fruitfully critiqued---Monica Bell (\protect\hyperlink{ref-Bell2017}{2017}) points to legal estrangement and structural exclusion in order to shift the theoretical focus from perceptions of unfairness to substantively unjust practices---the empirical literature on legitimacy is useful to demonstrate how police practices can alienate those subjected to them. The alienating effect of criminal legal exposure has also been tied to institutional avoidance (see, for instance, Brayne \protect\hyperlink{ref-Brayne2014}{2014}). .

@@

In order to ascertain whether fines and fees practices affect voter turnout, we exploit data from the 2012 and 2017 Census of Governments (COG) and national voter file data. The COG, a project of the US Census Bureau, collects budget data for all local governments every five years, and the most recent survey was in 2017. The COG asks cities how much revenue they collect from ``penalties imposed for violation of law; civil penalties (e.g., for violating court orders); court fees if levied upon conviction of a crime or violation\ldots{} and forfeits of deposits held for performance guarantees or against loss or damage (such as forfeited bail and collateral).'' As indicated by previous researchers, these data serve as a useful proxy for the prevalence of police ticketing (Goldstein, Sances, and You \protect\hyperlink{ref-Goldstein2020}{2020}; Sances and You \protect\hyperlink{ref-Sances2017}{2017}).

Unfortunately, the COG data does not include voter turnout, and turnout data is not generally available at this level. Therefore, in order to determine municipality-level turnout, we employ a national, geocoded voter file. This voter file is provided by L2 Political, and includes a host of information about each voter, including age, party affiliation, racial estimates, and historical turnout. By using the latitude and longitude of every registered voter in the country and shapefiles provided by the Census Bureau, we map each voter to their home municipality using the same definitions as the COG.\footnote{Although the registered voter file includes the name of the city in which the voter is registered, these do not always map neatly to the municipalities defined in the COG data. The use of these shapefiles ensures an apples-to-apples comparison.} We then aggregate these individual-level records up to the municipality level to determine each municipality's turnout.

Because the L2 national voter file includes estimates of each voter's racioethnic identity, we can also use these data to determine the \emph{racial turnout gap} in each municipality. These records are just estimates, and may be imprecise at the individual level. However, we assume that this presents less of a problem when aggregating to the municipality-level. In order to control for potential confounders, we also control for a host of other information from the 5-year American Community Survey and the COG.

@@

Figure \ref{fig:marg1} makes clear that higher fees and fines collections lead to lower turnout. The left-hand panel shows the simple bivariate relationship between turnout among registered voters in 2018 and fees and fines collection. The right-hand panel shows the bivariate relationship between the 2018 white-Black turnout gap and fines collections. The turnout gap is calculated by dividing the number of ballots cast by each racial group according to L2 by the estimated 2018 citizen voting age population (CVAP) from the Census Bureau to account for possible registration differences between groups.

@@

Although Figure \ref{fig:marg1} shows that there is a relationship between collections and turnout and the turnout gap, the simple scatter plots cannot demonstrate a causal relationship. To investigate whether there is such a relationship, we employ two-way fixed effect models. In these models, each municipality is observed twice: their first observation includes their (logged + 1) fees and fines per capita from the 2012 COG, their turnout and turnout gap in 2014, and their sociodemographic characteristics from the ACS 5-year estimates from 2012. The second observation includes their (logged + 1) fees and fines from the 2017 COG, their 2018 turnout (gap), and their 2018 5-year ACS demographics. In each case, turnout is measured only among the individuals who were registered in a given municipality in both 2014 and 2018 to control for changes in registration rates, and in Model 2 the turnout gap is calculated using CVAP. Table \ref{tab:tab1} presents the results of these models.

\begin{singlespace}
\input{"../temp/2wfe_reg_clean.tex"}
\end{singlespace}

@@

\hypertarget{refs}{}
\begin{cslreferences}
\leavevmode\hypertarget{ref-Bell2017}{}%
Bell, Monica C. 2017. ``Police Reform and the Dismantling of Legal Estrangement.'' \emph{The Yale Law Journal} 126 (7): 2054--2150. \url{http://www.jstor.org/stable/45222555}.

\leavevmode\hypertarget{ref-Brayne2014}{}%
Brayne, Sarah. 2014. ``Surveillance and System Avoidance: Criminal Justice Contact and Institutional Attachment.'' \emph{American Sociological Review} 79 (3): 367--91. \url{https://doi.org/10.1177/0003122414530398}.

\leavevmode\hypertarget{ref-CarpenterII2019}{}%
Carpenter II, Dick M, Kyle Sweetland, and Jennifer McDonald. 2019. ``The Price of Taxation by Citation: Case Studies of Three Georgia Cities That Rely Heavily on Fines and Fees.'' Institute for Justice.

@@

\leavevmode\hypertarget{ref-Desmond2016}{}%
Desmond, Matthew, Andrew V. Papachristos, and David S. Kirk. 2016. ``Police Violence and Citizen Crime Reporting in the Black Community.'' \emph{American Sociological Review} 81 (5): 857--76. \url{https://doi.org/10.1177/0003122416663494}.

@@

@@

\leavevmode\hypertarget{ref-Tyler2014}{}%
Tyler, Tom R., Jeffrey Fagan, and Amanda Geller. 2014. ``Street Stops and Police Legitimacy: Teachable Moments in Young Urban Men's Legal Socialization.'' \emph{Journal of Empirical Legal Studies} 11 (4): 751--85. \url{https://doi.org/10.1111/jels.12055}.

@@
